Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-06-17 doc/MS_v3.Rmd
more work on Intro

---
title: 'Sex-Specific Evolution of the Meiotic Recombination Rate'
date: '`r Sys.Date()`'
author: "April L. Peterson, Bret Payseur"
output:
  html_document:
  pdf_document:
    includes:
      in_header: header.tex
  word_document:
always_allow_html: yes
bibliography: refs_Main_Report.bib
csl: elife.csl
---

@@

Although meiotic recombination is required for successful gametogenesis in most species that reproduce sexually, the rate of crossing over varies among individuals. **define genome-wide recombination rate**.

Differences in recombination rate between females and males are perhaps the most striking form of this variation. Existing data fail to directly address the extent to which recombination experiences similar evolutionary pressures. To fill this gap, we measured meiotic recombination in both sexes for a panel of house mouse cross three subspecies. Using inbred strains and single cell immunohistochemistry allowed us to place sex-specific observations within the same genetic background and meiotic context. Our results indicate highly discordant evolutionary patterns in the two sexes. Whereas male recombination rates show evidence of rapid evolution over short evolutionary timescales, female recombination rates measured in the same strains are mostly static. These results strongly indicate that house mouse has two genome wide recombination rates which display distinct evolutionary trajectories.

@@

### 0.   Importance of recombination rate

Meiosis can be reduced to the expression (2n -> 4n -> 2n -> 1n) which reflects the process diploid genome producing haploid cell products. In most species the meiotic program relies on crossovers to ensure the correct separation of chromosomes at the reductional division (4n to 2n) [Cahoon]. The total number of COs per cell (at the 4n stage) is equal to the genome wide recombination rate (gwRR). Additioanlly, the gwRR can regulate a populations’ response to selection, and determine the fate of novel mutations by transferring beneficial mutations onto novel genetic backgrounds or by breaking linkage of negative mutations from beneficial genetic backgrounds (cite).  This process shapes the genomic patterns of genetic variation, with high recombination areas of genome having more nucleotide variation while areas of low recombination have decreased genetic variation [@begun1992; @nachman2012]. **Because of it's role in facilitating chromosomes disjunction, the genome wide recombination rate can be thought of as a cell-based metric which is tightly connected to organisms fitness and fertility.**

### 1. Understanding levels of variation in gwRR

The process of chromosome segregation is integral to the process of producing haploid gametes and can be considered a constraint in regards to the evolution of the genome-wide recombination rate.  Ensuring at least one crossover per chrm while minimizing the rates of DNA damage and ectopic recombination (cite) are considered to be the lower and upper thresholds for a genome-wide recombiantion rate [@nagaoka2012]. The number of haploid pairs of chromosomes is strong predictor for genome wide recombination rate across large divergence scales [@ottoPaysuer2019; @stapley_variation_2017]. Yet, within these thresholds the genome wide recombination rate still varies across species and individuals. Sex is one of the most notable axes of this form of variation. 
(examples, Humans 1.7 female to male
alligators female to male ratio is X, temperature dependent sex determination)

<2.	Gap is due to Lack of comparable measures>
An understanding of how sex shapes the evolution of recombination cannot be achieved with available data. Comprehensive comparisons of female and male recombination rates usually come from outbred populations humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], dog,  cattle [@ma2015_cattle; @Shen2018_cattle], (other sheep?) sheep [@johnston2016_soay], and mouse (CC cite) in which the role of sex is confounded with the contributions of genetic variation. 
(other gaps in data )

<3.	Open Q, no first principle>
Under anisogamy where gametogenesis is modified to produce distinct gametes (cite Gorelick?), the meiotic program is the same and there are no first principles which would predict the evolution of sexually dimorphic recombination rates. Yet heterochiasmy is commonly observed in dioecious species, suggesting that other meiotic traits which distinguish the gametes and their meiotic programs, for example symmetrical vs asymmetrical cell division, may impose selection for sexually dimorphic recombination rates. Although it is clear that the relationship between female and male recombination rates can differ among species, comparisons between and within closely related species are missing. Direct contrasts between females and males across a common, diverse set of genomic backgrounds would reveal whether the recombination rate evolves differently in the sexes.

<4.	Additional metrics (more nuanced)>
Initial focus/evidence on heterochiasmy was primarily focused on which sex had more recombination – the field has advanced since it’s first documentation (cite morgan 1914). Recently two specific features have been highlighted as conserved patterns of heterochiasmy: @CahoonLibuda2019 show that in species with meiotic chromatin (SC axis length) quantified for both sexes, longer chromosome axis generally has higher recombination rates. 
In @sardell_sex_2020, a survey of 51 species shows there are conserved broad scale recombination landscape differences. Generally males have telomere-bias crossover placement and females have more uniform placement.

These observations raise the question of how conservation in sexually dimorphic patterns are maintained with the gwRR and it’s decomposed traits.

@@

Because we know the meiotic programs, we have expectations for where the thresholds(constraints) of the gwRR 
the rate will fall. First, the lower threshold is defined as an obligate crossover per homolog pair which promotes proper chromosome segregation [@nagaoka2012]. second (while there is less evidence,), the upper threshold maybe regulated/shaped/affected due to rates of ectopic recombination or the programmed DNA damage preceding crossovers act as constraints to the upper threshold of meiotic recombination (cite). Yet, Within these thresholds the genome wide recombination rate still varies across species and indiviuals. 
which is the subject of an open question in the field.  **the gwRR is the best metric for since both of these constraints are thought to act at the cell wide level.**

**Sex is one of the most notable axes along which individuals vary**. Sexual dimorphism in recombination rates is called heterochiasmy [@lenormand2003]. An understanding of how sex shapes the evolution of recombination cannot be achieved with available data. Comprehensive comparisons of female and male recombination rates usually come from outbred populations humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], dog,  cattle [@ma2015_cattle; @Shen2018_cattle], (other sheep) sheep [@johnston2016_soay], and mouse (CC cite) in which the role of sex is confounded with the contributions of genetic variation. Under anisogamy where a species has distinct gametes, there is no first principle which would predict the evolution of sexually dimorphic recombination rates. Yet heterochiasmy, is commonly observed in dioecious species, suggesting that other meiotic traits which distinguish the gametes and their meiotic programs, for example symmetrical vs asymmetrical cell division, may impose selection for sexually dimorphic recombination rates. Although it is clear that the relationship between female and male recombination rates can differ among species, comparisons between and within closely related species are missing. Direct contrasts between females and males across a common, diverse set of genomic backgrounds would reveal whether the recombination rate evolves differently in the sexes.

<It’s important to note, – that the genome wide recombination rate can be decomposed – and connected to related traits (which also show conserved patterns of sexual dimorphism>

**understanding heterochiasmy and sexual dimorphism in meiosis have advanced since first being documented (morgan 1914) -- more than just the direction -- more evidence is accumulating that there are conserved patterns of sexual dimorphism in meiotic recombinatio - related to recombination rate**. Recently two specific features have been highlighted as conserved patterns of heterochiasmy:

@CahoonLibuda2019 show In species with meiotic chromatin (SC axis length) quantified for both sexes, longer chromosome axis generally has higher recombination rates. 
@sardell_sex_2020, A survey of 51 species shows there are conserved broad scale recombination landscape differences. Generally males have telomere-bias crossover placement and females have more uniform placement.

These observations raise the question of conservation in sexually patterns are maintained with the gwRR and it’s decomposed traits

**remember these points kinda distract from gwRR aspect**

### 2.  The House Mouse is a great model for uncovering evolutionary patterns at a short timescale

The House mouse species complex arose from a recent radiation providing an opportunity to interrogate natural variation at short evolutionary scales, potentially narrowing the mutational space for identifying variants which would cause gwRR variation. Add divergence ranges. divergence spanning 0.5 mya to 5 mya  (cite / Geraldes et al 2008 and Geraldes et al 2010).

Classical lab strains of mice have generated extensive knowledge central to meiosis and outcomes on recombination through extensive studies on the genes involved in the meiotic recombination pathway (cite) and previous crosses for understanding the genetic architecture of recombination rate variation @dumont2011house; @murdoch2010; @Wang2017island].
Additionally due to their global distribution allows us to compare samples of natural diversity from a broad geographic range.  (wild derived inbred strains – will capture genetic diversity not represented in classical lab strains.)

@@

While there are strains with Robertsonian translocations, in house mouse and related murid species it is possible to assemble sets of strains with identical karyotypes, 20 pairs of acrocentric chromosomes.

House mouse is well suited for single cell cytology approaches (due to large cell sizes). The single cell level allows integration of data at a closer connection to the molecular pathway and meiotic program. In addition to the quantification of gwRR. This study quantifies precursors to crossovers, double strand breaks (DSBs), and meiotic chromosome morphology across stages of meiosis which would not be accessible to with genetic linkage data and crosses.

@@

The House mouse complex arose from a recent radiation providing an opportunity to interrogate natural variation at short evolutionary scales, potentially narrowing the mutational space for identifying variants which would cause gwRR variation. Add divergence ranges. divergence spanning 0.5 mya to 5 mya.

Classical lab strains of mice have generated extensive knowledge central to meiosis and outcomes on recombination through extensive studies on the genes involved in the meiotic recombination pathway (cite) and previous crosses for understanding the genetic architecture of recombination rate variation @dumont2011house; @murdoch2010; @Wang2017island].

**Wild derived inbred strains generate the best comparison of females and males, besides the sex chromosomes, the mouse for each genome are almost identical**. Inbred strains enable one of the most direct comparison of both female and male --  versions of meiosis --enable use to conclude that sex is a primary factor --

Additionally due to their global distribution allows us to compare samples of natural diversity from a broad geographic range.  (wild derived inbred strains – will capture genetic diversity not represented in classical lab strains.)

While there are strains with Robertsonian translocations, in house mouse and related murid species it is possible to assemble sets of strains with identical karyotypes, 20 pairs of acrocentric chromosomes.

House mouse is well suited for single cell cytology approaches. The single cell level allows integration of data at a closer connection to the molecular pathway and meiotic program. In addition to the quantification of gwRR. This study quantifies precursors to crossovers, double strand breaks (DSBs), and meiotic chromosome morphology across stages of meiosis which would not be accessible to with genetic linkage data and crosses.

### 3. What we accomplished in this paper

Main messages: these results are centered/focused on a unique evo comparison 
Advantages / importance of using multiple strains from and multiple subspecies
-importance / advantage of sampling for multiple geographic locations.
Chromosome morphology / general recombination landscape
-results for general landscape – conserved sex differences  and

---------------
OUR results for genome wide recombination rate (variation) we observed highly distinct evolutionary trajectories between males and females. This is due to rapid male specific evolution in the gwRR from two strains in musculus and 1 strain in molossinus. (this pattern of rapid evolution was not observed in females). We observed up to a 30% difference (translating into approximately ~7 more crossovers per cell), a surprising amount considering the short evolutionary time scale.

We quantified the general recombination landscape because patterns in the recombination landscape are important to document because they can shed light on constraints and regulation of the processes ( could end up leading to variation in the genome wide recombination rates. Our results confirm (well established) previously reported patterns for sex differences in the general recombination landscape – and add by demonstrating that these patterns hold even in the case of genome more crossovers, but general landscape patterns still maintain sex-specific patterns.

Conserved patterns of the recombination landscape of the single chromosome level – reflect constraints  At the single chromosome level – (crossover number is conserved – suggesting processes regulating the rec landscape at this scale might face similar constraints

@@

For the first time in many of these strains, We report a rare, direct, evolutionary comparison of recombination rate in females and males for this short evolutionary scale. We choose Mus musculus because of the available strains across subspecies. Our design of using multiple inbred strains within subspecies  and multiple subspecies across the species --  (can provide information anaglogous to polymorphism within and divergence within subspecies of the house mouse. Our results indicate rapid male specific evolution of gwRR occurred in a subset of house mouse lines. We observed up to a 30% difference (translating into approximately ~7 more crossovers per cell), a surprising amount considering the short evolutionary time scale.  (found previously for PWK (forjt) (PWD beth) – but mostly measures from males.

@@

We quantified the general recombination landscape because patterns in the recombination landscape are important to document because they can shed light on constraints and regulation of the processes ( could end up leading to variation in the genome wide recombination rates. Our results confirm (well established) previously reported patterns for sex differences in the general recombination landscape – and add by demonstrating that these patterns hold even in the case of genome more crossovers, but general landscape patterns still maintain sex-specific patterns.

>>>>

---
title: 'Sex-Specific Evolution of the Meiotic Recombination Rate'
date: '`r Sys.Date()`'
author: "April L. Peterson, Bret Payseur"
output:
  word_document:
  pdf_document:
    includes:
      in_header: header.tex
  html_document:
always_allow_html: yes
bibliography: refs_Main_Report.bib
csl: elife.csl
---

@@

Although meiotic recombination is required for successful gametogenesis in most species that reproduce sexually, the rate of crossing over varies among individuals. The number of crossovers per meiotic cell is the  genome-wide recombination rate. Differences in recombination rate between females and males are perhaps the most striking form of this variation. Existing data fail to directly address the extent to which recombination experiences similar evolutionary pressures. To fill this gap, we measured meiotic recombination in both sexes for a panel of house mouse cross three subspecies. Using inbred strains and single cell immunohistochemistry allowed us to place sex-specific observations within the same genetic background and meiotic context. Our results indicate highly discordant evolutionary patterns in the two sexes. Whereas male recombination rates show evidence of rapid evolution over short timescales, female recombination rates measured in the same strains are mostly static. These results strongly indicate that house mouse has two, one for each sex, genome wide recombination rates which display distinct evolutionary trajectories.

@@

<!--### 0.   Importance of recombination rate-->

  Meiosis can be reduced to the expression (2n -> 4n -> 2n -> 1n) which reflects the process of a diploid genome producing haploid gametes. In most species the meiotic program relies on crossovers to ensure the correct separation of chromosomes at the reductional division (4n to 2n) [@CahoonLibuda2019]. The total number of COs at this stage of meiosis is equal to the genome wide recombination rate (gwRR). In addition to this functional purpose, the gwRR can regulate a populations’ response to selection, and determine the fate of novel mutations by transferring beneficial mutations onto novel genetic backgrounds or by breaking linkage of negative mutations from beneficial genetic backgrounds (Charlesworth 1993 cite).  This process shapes the genomic patterns of genetic variation, with high recombination areas of genome having more nucleotide variation while areas of low recombination have decreased genetic variation [@begun1992; @nachman2012].

<!-- ### 1. Understanding levels of variation in gwRR -->

  **Due to it's role in facilitating chromosomes disjunction, the genome wide recombination rate can be thought of as a cell-based metric which is tightly connected to organisms fitness and fertility.**  Ensuring at least one crossover per chrm while minimizing the rates of DNA damage and ectopic recombination are thought to regulate the lower and upper thresholds for a genome-wide recombiantion rate [@nagaoka2012] (Inoue). Yet, within these thresholds the genome wide recombination rate still varies across species and individuals. Sex is one of the most notable axes of this form of variation.

<!-- (examples, Humans 1.7 female to male
alligators female to male ratio is X, temperature dependent sex determination) -->

<!-- The number of haploid pairs of chromosomes is strong predictor for genome wide recombination rate across large divergence scales [@ottoPaysuer2019; @stapley_variation_2017]. -->

<!-- 2.Gap is due to Lack of comparable measures -->

An understanding of how sex shapes the evolution of recombination cannot be achieved with available data. Comprehensive comparisons of female and male recombination rates usually come from outbred populations humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], dog,  cattle [@ma2015_cattle; @Shen2018_cattle], (other sheep?) sheep [@johnston2016_soay], and mouse (CC cite) in which the role of sex is confounded with the contributions of genetic variation.

<!-- 3.	Open Q, no first principle -->

Even under anisogamy, where gametogenesis is modified to produce distinct gametes across sexes, the meiotic program is the same. There are no first principles which would predict the evolution of sexually dimorphic recombination rates. Yet heterochiasmy is commonly observed in dioecious species, suggesting that other meiotic traits which distinguish the gametes and their meiotic programs, for example symmetrical vs asymmetrical cell division, may impose selection for sexually dimorphic recombination rates. Although it is clear that the relationship between female and male recombination rates can differ among species, comparisons between and within closely related species are missing. Direct contrasts between females and males across a common, diverse set of genomic backgrounds would reveal whether the recombination rate evolves differently in the sexes.

<!-- 4.	Additional metrics (more nuanced) -->  
  Heterochiasimy was defined as sexual dimorphism in the total counts measures of recombination (Lenormand 2003). It's important to note that the genome-wide recombiantion rate can be decomposed into single chromosome patterns, refered to as recombination landscapes. Recently two specific features have been highlighted as conserved patterns related heterochiasmy, specifically the recombination landscape. In a review of data from mouse, celegans and worms with meiotic chromosome morphilogy (axis-loop structure) quantified for both sexes, @CahoonLibuda2019 find that the sex with longer chromosome axis generally has higher recombination rates. A survey of 51 species shows there are conserved broad scale recombination landscape differences. Generally males have telomere-bias crossover placement and females have more uniform placement [@sardell_sex_2020]. These observations raise the question of how conservation in sexually dimorphic patterns are maintained with the gwRR and it’s decomposed traits.

@@

<!-- ### 2.  The House Mouse is a great model for uncovering evolutionary patterns at a short timescale -->

  The house mouse species, (_Mus musculus_), complex arose from a recent radiation (0.5 mya to 5 mya) (@geraldes2011; @geraldes2008) providing an opportunity to interrogate natural variation at short evolutionary scales. This feature narrows the mutational space for identifying variants which would cause gwRR variation.  Classical lab strains of mice have generated extensive knowledge central to meiosis and outcomes on recombination through extensive studies on the genes involved in the meiotic recombination pathway **(cite)** and previous crosses for understanding the genetic architecture of recombination rate variation [@dumont2011; @murdoch2010; @Wang2017island]. Additionally due to their global distribution allows us to compare samples of natural diversity from a broad geographic range.

@@

  House mouse is well suited for single cell cytology approaches (due to large cell sizes). The single cell level allows integration of data at a closer connection to the molecular pathway and meiotic program. (Allows us to compare counts from cells at is the same meiotic (molecular pathway context)).

@@

<!-- ### 3. What we accomplished in this paper -->

<!-- Main messages: these results are centered/focused on a unique evo comparison 
Advantages / importance of using multiple strains from and multiple subspecies
-importance / advantage of sampling for multiple geographic locations.
Chromosome morphology / general recombination landscape
-results for general landscape 
– conserved sex differences  -->

---------------

  (we quantify gwRR from both sexes, using house mouse inbred strains as a common genetic background). We observed highly distinct evolutionary trajectories between males and females. This is due to rapid male specific evolution in the gwRR from two strains in musculus and 1 strain in molossinus. (this pattern of rapid evolution was not observed in females). We observed up to a 30% difference (translating into approximately ~7 more crossovers per cell), a surprising amount considering the short evolutionary time scale.

<!-- we decompsed the gwRR into single chromosome features to access data which  reflects the non-random processes of crossover formation on individual chromosomes.-->

  The recombination landscape for single chromosome scale can be general summarized by i) spacing of crossovers and ii) the linear area of synapsed homologous chromosomes where crossovers can form. We quantified the general recombination landscape because patterns in the recombination landscape are important to document because they can shed light on constraints and regulation of the processes ( could end up leading to variation in the genome wide recombination rates. Our results confirm (well established) previously reported patterns for sex differences in the general recombination landscape – and add by demonstrating that these patterns hold even in the case of genome more crossovers, but general landscape patterns still maintain sex-specific patterns. Conserved patterns of the recombination landscape of the single chromosome level – reflect constraints  At the single chromosome level – (crossover number is conserved – suggesting processes regulating the rec landscape at this scale might face similar constraints

@@

<!-- For the first time in many of these strains, We report a rare, direct, evolutionary comparison of recombination rate in females and males for this short evolutionary scale. We choose Mus musculus because of the available strains across subspecies. Our design of using multiple inbred strains within subspecies  and multiple subspecies across the species --  (can provide information anaglogous to polymorphism within and divergence within subspecies of the house mouse. Our results indicate rapid male specific evolution of gwRR occurred in a subset of house mouse lines. We observed up to a 30% difference (translating into approximately ~7 more crossovers per cell), a surprising amount considering the short evolutionary time scale.  (found previously for PWK (forjt) (PWD beth) – but mostly measures from males.

@@

We quantified the general recombination landscape because patterns in the recombination landscape are important to document because they can shed light on constraints and regulation of the processes ( could end up leading to variation in the genome wide recombination rates. Our results confirm (well established) previously reported patterns for sex differences in the general recombination landscape – and add by demonstrating that these patterns hold even in the case of genome more crossovers, but general landscape patterns still maintain sex-specific patterns. -->
